---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Synthetic Control with Differencing (SCD)

by Chun Pang Chow, Ratzanyel Rincón, and Kyungchul Song.

## Description

The `scd` package implements the SCD method in Rincón and Song (2025) to estimate the average treatment effects on the treated (ATT) and their asymptotically valid confidence intervals in settings with one treated group and multiple untreated donor groups over short panels or repeated cross-sections. Identification relies on a Stable Matching Condition (SMC), offering a robust alternative to difference-in-differences (DID) when the parallel trend assumption is unlikely to hold.

## Installation

You can install the `scd` package in `R` from Github with:

```{r, eval=FALSE}
# Install devtools using: install.packages("devtools")
devtools::install_github("ratzanyelrincon/scd")
```

## Quick Example

In this example we analyze the impact of the 2007 Legal Arizona Workers Act (LAWA) on weekly earnings of individuals living in Arizona. The built-in dataset is a 5% random sample of the U.S. Current Population Survey (CPS) Basic Monthly Files, between January 2003 and December 2009. The data were obtained using the `cpsR` package in `R` and include individual-level observations with log weekly earnings (*wklyearn*), a treatment dummy (*D*), time variables (*period* and *period_label*), state identifiers (*statefip* and *state_name*), and sampling weights (*weight*). Observations with non-positive weekly earnings are dropped. 

Following Bohn et al. (2014), we consider Arizona (*statefip* = 4) as treated after the passage of LAWA in July 2007 and include in the donor pool the other 46 states that did not implement any similar regulation during the period of analysis. The data structure is repeated cross-sections with 700 observations per month on average. You can read the data in `R` and have a glimpse of it as follows:

```{r}
library(dplyr)
library(scd)

# Read data
data <- CPSdta

# Take a look at the first rows of the data
head(data)
```

To prepare the data for estimating the target parameters, run:

```{r}
t_labels <- sort(unique(data$period_label))

treated <- 4

donors <- data %>%
  distinct(statefip, state_name) %>%
  filter(statefip != treated) %>%
  arrange(statefip)

guntreated <- donors %>%
  pull(statefip)

guntreated_labels <- donors %>%
  pull(state_name)
```

To estimate the average treatment effects using the DID differencing parameter, use the `att_scd` function as follows:

```{r}
result <- att_scd(data = data,
                  yname = "wklyearn",
                  dname = "D",
                  tname = "period",
                  gname = "statefip",
                  gtreated = treated,
                  guntreated = guntreated,
                  t_labels = t_labels,
                  guntreated_labels = guntreated_labels,
                  wname = "weight",
                  lambda = "DID",
                  data_type = "RC",
                  post_treatment_only = FALSE)
```


`att_scd` returns a list of objects including the donor pool weights, the synthetic counterfactual for *wklyearn*, and the ATT estimates along with their corresponding 95% confidence intervals for each period. 

To control how within-group time differencing is handled in `att_scd`, users can adjust the parameter *lambda*. The available options are: "DID", which differences outcomes relative to the last pre-treatment period; "Unif", which differences outcomes relative to the average of all pre-treatment periods; and "SC", which applies no time differencing to the outcome. A custom vector *lambda* can be provided as well.   

We can then create an event study plot using the function `plot_att_scd`:

```{r}
library(ggplot2)

plot_att_scd(result)
```

The blue and red points display the estimated average treatment effects on the treated in the pre- and post-treatment periods, respectively, together with their corresponding 95% confidence intervals. Under SMC, the post-LAWA estimates can be interpreted as the causal effect of the policy on the log weekly earnings in Arizona. Similar to the usual pre-trend test found in empirical studies using DID, the non-statistically significant estimates of the treatment effects during the pre-treatment period provide evidence in favour of the stability assumption of matching weights in SCD.

## References

Bohn, S., Lofstrom, M., and Raphael, S. (2014). Did the 2007 Legal Arizona Workers Act Reduce the State’s Unauthorized Immigrant Population? The Review of Economics and Statistics, 96(2), 258–269. <http://www.jstor.org/stable/43554929>

Callaway, B., and Sant’Anna, P. H. C. (2021). Difference-in-Differences with Multiple Time Periods. Journal of Econometrics, 225(2), 200–230. <https://doi.org/10.1016/j.jeconom.2020.12.001>

Rincón, R., and Song, K. (2025). Causal Inference with Groupwise Matching. *arXiv preprint*, arXiv:2510.26106 <https://arxiv.org/abs/2510.26106>

The technical appendix can be downloaded [here](https://raw.githubusercontent.com/ratzanyelrincon/scd/master/Technical_Appendix.pdf).
