---
title: "Introduction to Synthetic Control with Differencing"
author: "Chun Pang Chow, Ratzanyel Rincón, and Kyungchul Song"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to Synthetic Control with Differencing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

The **scd** package implements the Synthetic Control with Differencing (SCD) method for estimating average treatment effects on the treated (ATT) in settings with one treated group and multiple untreated donor groups observed over time. The method, proposed in Rincón and Song (2025), provides asymptotically valid uniform confidence intervals under a **Generalized Matching Condition (GMC)**, which relaxes the parallel trends assumption required by standard difference-in-differences.

### When to Use SCD

SCD is particularly useful when:

1. **Parallel trends may not hold** — The treated unit may have had different trends than donors even absent treatment
2. **You have repeated cross-sections or short panels** — Individual-level data need not be tracked over time
3. **Multiple large donor groups are available** — SCD leverages group-level variation rather than aggregate time series
4. **You want uniform inference** — Confidence bands are valid simultaneously across all post-treatment periods (or all periods if `post_treatment_only = FALSE`)

## Installation and Setup

```{r install, eval = FALSE}
# Development version from GitHub
# install.packages("devtools")
devtools::install_github("whereisalexchow/scd")
```

```{r load}
library(scd)
library(dplyr)
library(ggplot2)
```

## The `att_scd()` Function

The main function `att_scd()` estimates period-specific ATT and constructs uniform confidence intervals.

### Function Signature

```{r signature, eval = FALSE}
att_scd(
  data,                          # Data frame with panel/cross-section data
  yname,                         # Name of outcome variable
  dname,                         # Name of treatment indicator (0/1)
  tname,                         # Name of time variable (integer 1,...,T)
  gname,                         # Name of group variable
  gtreated,                      # Value identifying treated group
  guntreated,                    # Vector of donor group identifiers
  t_labels,                      # Labels for time periods
  guntreated_labels,             # Labels for donor groups
  idname = NULL,                 # Individual ID variable (required for panel)
  wname = NULL,                  # Optional: survey weight variable
  lambda = "DID",                # Differencing scheme
  data_type = "auto",            # "auto", "panel", or "RC"
  L = 0,                         # Post-treatment periods (0 = all available)
  alpha = 0.05,                  # Significance level
  kappa = 0.005,                 # Bonferroni adjustment parameter
  length_grid_w = 500,           # Number of weight grid points
  n_wpoints_per_edge = 20,       # Points per simplex edge
  tolerance = 1e-06,             # Numerical tolerance for binding constraints
  parallel = TRUE,               # Use parallel computation
  post_treatment_only = TRUE     # Compute CIs only for post-treatment periods
)
```

### Key Arguments

#### Data Structure (`data_type`)

| Value | Description |
|-------|-------------|
| `"auto"` | Auto-detect: panel if `idname` provided and each individual observed in every period; RC otherwise |
| `"panel"` | Balanced panel. Requires `idname`. Uses panel variance estimator (accounts for within-individual correlation) |
| `"RC"` | Repeated cross-sections. Uses RC variance estimator (exploits cross-period independence for sharper inference) |

#### Differencing Scheme (`lambda`)

The `lambda` argument controls how pre-treatment outcomes are used to construct the baseline:

| Value | Formula | Description |
|-------|---------|-------------|
| `"DID"` | $\lambda = (0, \ldots, 0, 1)$ | Uses only the last pre-period (standard DiD) |
| `"Unif"` | $\lambda = (1/T_0, \ldots, 1/T_0)$ | Uniform average over all pre-periods |
| `"SC"` | $\lambda = (0, \ldots, 0)$ | No differencing (pure synthetic control) |
| Custom | User-specified vector in $\Delta_{T_0-1}$ | Domain-specific weighting |

**Guidance:** Use `"Unif"` when pre-treatment outcomes are relatively stable across time. Use `"DID"` when levels shift but slopes are stable. Use a custom vector when you have domain knowledge about which pre-periods are most informative.

#### Pre-Treatment CIs (`post_treatment_only`)

By default, `post_treatment_only = TRUE` and confidence intervals are computed only for post-treatment periods. Pre-treatment periods still receive point estimates ($\hat{\theta}_t$) but their CI columns will be `NA`. Set `post_treatment_only = FALSE` to compute CIs for all periods; this is useful for diagnostic pre-trend checks but increases computation time.

## Example: Arizona Immigration Enforcement

We analyze the effect of Arizona's Legal Arizona Workers Act (LAWA) on Hispanic workers' earnings using the included `CPSdta` dataset.

### Data Preparation

```{r data-prep}
# Load the CPS data
data("CPSdta")

# Examine the data structure
glimpse(CPSdta)

# The data contains:
# - wklyearn: log weekly earnings (outcome)
# - D: treatment indicator (1 = Arizona post-July 2007)
# - period: time period index
# - statefip: state FIPS code (group identifier)
# - weight: survey weights
```

```{r setup-groups}
# Define the treated group
treated <- 4  # Arizona's FIPS code

# Define donor groups (all other states)
donors <- CPSdta %>%
  distinct(statefip, state_name) %>%
  filter(statefip != treated) %>%
  arrange(statefip)

# Extract identifiers and labels
guntreated <- donors$statefip
guntreated_labels <- donors$state_name
t_labels <- sort(unique(CPSdta$period_label))
```

### Running the Estimation

```{r estimate}
# Estimate ATT using uniform differencing with CIs for all periods
result <- att_scd(
  data = CPSdta,
  yname = "wklyearn",
  dname = "D",
  tname = "period",
  gname = "statefip",
  gtreated = treated,
  guntreated = guntreated,
  t_labels = t_labels,
  guntreated_labels = guntreated_labels,
  wname = "weight",
  lambda = "Unif",
  data_type = "RC",
  alpha = 0.05,
  parallel = TRUE,
  post_treatment_only = FALSE  # Include pre-treatment CIs for diagnostics
)
```

### Examining Results

```{r results}
# Estimated weights on donor groups
head(result$donorpool_weights)

# Period-specific ATT estimates with confidence intervals
result$target_parameter

# Summary statistics
cat("Data type:", result$data_type, "\n")
cat("Sample size:", result$n, "\n")
cat("Number of donors:", result$K, "\n")
cat("Pre-treatment periods:", result$T_0, "\n")
cat("Post-treatment periods:", result$T_1, "\n")
cat("Computation time:", result$time_elapsed, "\n")
```

### Visualization

```{r plot, fig.width = 8, fig.height = 5}
# Create event study plot
plot_att_scd(result)

# With numeric x-axis instead of calendar dates
plot_att_scd(result, date_start = NULL)
```

The plot shows:

- **Blue points:** Pre-treatment ATT estimates (should be near zero under GMC)
- **Red points:** Post-treatment ATT estimates (causal effects)
- **Error bars:** 95% uniform confidence intervals

## Advanced Usage

### Comparing Differencing Schemes

```{r compare-lambda}
# DID-style differencing
result_did <- att_scd(
  data = CPSdta,
  yname = "wklyearn", dname = "D", tname = "period", gname = "statefip",
  gtreated = treated, guntreated = guntreated,
  t_labels = t_labels, guntreated_labels = guntreated_labels,
  wname = "weight",
  lambda = "DID",
  data_type = "RC"
)

# Uniform differencing
result_unif <- att_scd(
  data = CPSdta,
  yname = "wklyearn", dname = "D", tname = "period", gname = "statefip",
  gtreated = treated, guntreated = guntreated,
  t_labels = t_labels, guntreated_labels = guntreated_labels,
  wname = "weight",
  lambda = "Unif",
  data_type = "RC"
)

# Compare post-treatment confidence interval widths
ci_width <- function(r) {
  post <- r$target_parameter[(r$T_0 + 1):(r$T_0 + r$T_1), ]
  mean(post$uci_hat_theta_scd - post$lci_hat_theta_scd, na.rm = TRUE)
}

cat("Mean post-treatment CI width (DID):", round(ci_width(result_did), 3), "\n")
cat("Mean post-treatment CI width (Unif):", round(ci_width(result_unif), 3), "\n")
```

## Output Structure

The `att_scd()` function returns a list with the following components:

| Component | Description |
|-----------|-------------|
| `donorpool_weights` | Data frame with donor identifiers, labels, and estimated weights $\hat{w}$ |
| `target_parameter` | Data frame with `period`, `hat_theta_scd`, `lci_hat_theta_scd`, `uci_hat_theta_scd` |
| `synthetic_outcome` | Data frame with observed (`Y`) vs. synthetic (`Y_synth`) outcomes |
| `hat_sigma` | Estimated standard errors $\hat{\sigma}_t$ for each period |
| `n` | Total sample size (unique individuals for panel; sum of weights for RC) |
| `K` | Number of donor groups |
| `T_0` | Number of pre-treatment periods |
| `T_1` | Number of post-treatment periods |
| `T_star` | First treatment period |
| `L` | Number of post-treatment periods used |
| `lambda` | Differencing scheme used |
| `data_type` | `"panel"` or `"RC"` |
| `post_treatment_only` | Whether CIs were restricted to post-treatment |
| `conf_set_size` | Size of weight confidence set |
| `time_elapsed` | Computation time |

## Troubleshooting

### Common Issues

1. **"Error: hat_Gamma_pre not full rank"**
   - Cause: Collinearity among donor groups or too few pre-periods ($T_0 < K$)
   - Solution: Reduce number of donors or combine similar groups

2. **"Error: hat_V not positive definite"**
   - Cause: Near-singularity in variance estimation
   - Solution: Increase sample size or reduce number of donors

3. **Empty confidence set (Bonferroni)**
   - Cause: Grid search missed the confidence region
   - Solution: Increase `length_grid_w` and `n_wpoints_per_edge`

4. **All CIs are `NA` for pre-treatment periods**
   - Cause: Default `post_treatment_only = TRUE` skips pre-treatment CIs
   - Solution: Set `post_treatment_only = FALSE`

5. **Slow computation**
   - Solution: Set `parallel = TRUE`, use `post_treatment_only = TRUE` (default), or reduce `length_grid_w`

## References

- Rincón, R. and Song, K. (2025). "Causal Inference with Groupwise Matching." Working Paper, Vancouver School of Economics, UBC.
